### 2014-04-09 ApTranscriptome_TR.Rmd
Gene set enrichment analysis for expression type overall and each colony

**April 4, 2014**

@@

```{r temperature_response, eval=TRUE, echo=FALSE, results='asis'}
# table type of response
A22.type.table <- table(Ap.response.type[ , 'A22.type'])
Ar.type.table <- table(Ap.response.type[ , 'Ar.type'])

Ap.type.table <- rbind(A22.type.table, Ar.type.table)
rownames(Ap.type.table) <- c("A22", "Ar")

@@

pandoc.table(Ap.type.table, style="rmarkdown", caption = "Number of thermally-responsive transcripts with maximum expression at high, low, intermediate, both high and low (bimodal) temperatures or are not expressed for each colony.")

@@

```{r plotthermalbreadth, echo=FALSE, eval=TRUE)
# prep data for plotting
Ap.int.sd <- rbind(A22.int.sd, Ar.int.sd)

@@

```{r GO_contingency_test}
# GO 'response to stress' hits in responsive transcripts
responsive.transcripts[grep("GO:0006950", responsive.transcripts$GO.Biological.Process), list(Transcript, best.hit.to.nr)]

resp.stress.responsive.count <- nrow(responsive.transcripts[grep("GO:0006950", responsive.transcripts$GO.Biological.Process), list(Transcript, best.hit.to.nr)])
# GO 'response to stress' hits in all transcripts
resp.stress.all.count <- nrow(annotationtable[grep("GO:0006950", annotationtable$GO.Biological.Process), list(Sequence.Name, best.hit.to.nr)])

GO.stress.table <- matrix(rbind(resp.stress.responsive.count, nrow(responsive.transcripts) - resp.stress.responsive.count, resp.stress.all.count, nrow(annotationtable) - resp.stress.all.count), nrow = 2)

@@

There are also 4 heat shock related genes in the responsive transcripts, out of 130 total.

```{r grep_shock}
hsp_responsive <- responsive.transcripts[grep("shock", responsive.transcripts$best.hit.to.nr), list(Transcript, best.hit.to.nr)]
nrow(hsp_responsive)

@@

```{r gsea_all}
# create geneList. note that NA values cause problems with topGO
# so set any NA to 1 as need to retain for GO analysis
Ap.geneList <- RxNout$padj
Ap.geneList[which(is.na(Ap.geneList))] <- 1
stopifnot(length(which(is.na(Ap.geneList))) == 0)
names(Ap.geneList) <- RxNout$Transcript
str(Ap.geneList)

@@

# table results
Ap.BP.ResTable <- GenTable(Ap.BP.GOdata, parentchild = Ap.BP.resultParentChild, topNodes = 118)
#Ap.BP.ResTable
write.table(Ap.BP.ResTable, file = paste(resultsdir, "Ap_GO.BP_results.txt", sep=""), quote=FALSE, row.names=FALSE, sep = "\t")
pandoc.table(Ap.BP.ResTable)

@@

**2) High and low expressed transcripts**

Use `selectTranscript` function to select transcripts based on 'A22.exp.type'. As only significant transcripts have an 'A22.exp.type' assigned, this is a small subset of the above.

```{r A22.exp.type_function}
selectTranscript <- function(score) {
    return(score == 1)
}

# To select genes, add 'A22.exp.type' to RxN df
RxNout.t <- merge(RxNout, Ap.A22.exp.type, by="Transcript", all=TRUE)
length(which(!is.na(RxNout.t$A22.exp.type)))
```

```{r bim.gsea, echo=TRUE, eval=TRUE}
# create geneList
Ap.bim.geneList <- rep(0, length = nrow(RxNout.t))
Ap.bim.geneList[which(RxNout.t$A22.exp.type=="Bimodal")] <- 1
names(Ap.bim.geneList) <- RxNout.t$Transcript
str(Ap.bim.geneList)
table(Ap.bim.geneList)

# create topGOdata object
Ap.bim.BP.GOdata <- new("topGOdata",
                 description = "BP gene set analysis",
                 ontology = "BP",
                 allGenes = Ap.bim.geneList,
                 geneSel = selectTranscript,
                 nodeSize = 10,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO)

Ap.bim.BP.GOdata

# perform enrichment analysis using parentchild method
Ap.bim.BP.resultParentChild <- runTest(Ap.bim.BP.GOdata, statistic = 'fisher', algorithm = 'parentchild')
Ap.bim.BP.resultParentChild
```

Enriched gene sets for genes expressed at both high and low temperatures.

```{r, echo=FALSE, results="asis"}
Ap.bim.BP.ResTable <- GenTable(Ap.bim.BP.GOdata, parentchild = Ap.bim.BP.resultParentChild, topNodes = 52)
#Ap.bim.BP.ResTable
write.table(Ap.bim.BP.ResTable, file = paste(resultsdir, "Ap_bimodal_GO.BP_results.txt", sep=""), quote=FALSE, row.names=FALSE, sep = "\t")
pandoc.table(Ap.bim.BP.ResTable)
```

**3) Intermediate-temperature expressed transcripts**

```{r int.gsea, echo=TRUE, eval=TRUE}
# create geneList
Ap.int.geneList <- rep(0, length = nrow(RxNout.t))
Ap.int.geneList[which(RxNout.t$A22.exp.type=="Intermediate")] <- 1
names(Ap.int.geneList) <- RxNout.t$Transcript
str(Ap.int.geneList)
table(Ap.int.geneList)

# create topGOdata object
Ap.int.BP.GOdata <- new("topGOdata",
                 description = "BP gene set analysis",
                 ontology = "BP",
                 allGenes = Ap.int.geneList,
                 geneSel = selectTranscript,
                 nodeSize = 10,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO)

Ap.int.BP.GOdata

# perform enrichment analysis using parentchild method
Ap.int.BP.resultParentChild <- runTest(Ap.int.BP.GOdata, statistic = 'fisher', algorithm = 'parentchild')
Ap.int.BP.resultParentChild
```

Enriched gene sets for genes expressed at intermediate temperatures.

```{r, echo=FALSE, results="asis"}
Ap.int.BP.ResTable <- GenTable(Ap.int.BP.GOdata, parentchild = Ap.int.BP.resultParentChild, topNodes=16)
Ap.int.BP.ResTable
write.table(Ap.int.BP.ResTable, file = paste(resultsdir, "Ap_intermediate_GO.BP_results.txt", sep=""), quote=FALSE, row.names=FALSE, sep = "\t")
pandoc.table(Ap.int.BP.ResTable)
```

**4) High temperature expressed transcripts**

```{r hig.gsea, echo=TRUE, eval=TRUE}
# create geneList
Ap.hig.geneList <- rep(0, length = nrow(RxNout.t))
Ap.hig.geneList[which(RxNout.t$A22.exp.type=="High")] <- 1
names(Ap.hig.geneList) <- RxNout.t$Transcript
str(Ap.hig.geneList)
table(Ap.hig.geneList)

# create topGOdata object
Ap.hig.BP.GOdata <- new("topGOdata",
                 description = "BP gene set analysis",
                 ontology = "BP",
                 allGenes = Ap.hig.geneList,
                 geneSel = selectTranscript,
                 nodeSize = 10,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO)

Ap.hig.BP.GOdata

# perform enrichment analysis using parentchild method
Ap.hig.BP.resultParentChild <- runTest(Ap.hig.BP.GOdata, statistic = 'fisher', algorithm = 'parentchild')
Ap.hig.BP.resultParentChild
```

Enriched gene sets for genes expressed at both high temperatures only.

```{r, echo=FALSE, results="asis"}
Ap.hig.BP.ResTable <- GenTable(Ap.hig.BP.GOdata, parentchild = Ap.hig.BP.resultParentChild, topNodes=12)
#Ap.hig.BP.ResTable
write.table(Ap.hig.BP.ResTable, file = paste(resultsdir, "Ap_high_GO.BP_results.txt", sep=""), quote=FALSE, row.names=FALSE, sep = "\t")
pandoc.table(Ap.hig.BP.ResTable)
```

**5) Low temperature expressed transcripts**

```{r low.gsea, echo=TRUE, eval=TRUE}
# create geneList
Ap.low.geneList <- rep(0, length = nrow(RxNout.t))
Ap.low.geneList[which(RxNout.t$A22.exp.type=="Low")] <- 1
names(Ap.low.geneList) <- RxNout.t$Transcript
str(Ap.low.geneList)
table(Ap.low.geneList)

# create topGOdata object
Ap.low.BP.GOdata <- new("topGOdata",
                 description = "BP gene set analysis",
                 ontology = "BP",
                 allGenes = Ap.low.geneList,
                 geneSel = selectTranscript,
                 nodeSize = 10,
                 annot = annFUN.gene2GO,
                 gene2GO = geneID2GO)

Ap.low.BP.GOdata

# perform enrichment analysis using parentchild method
Ap.low.BP.resultParentChild <- runTest(Ap.low.BP.GOdata, statistic = 'fisher', algorithm = 'parentchild')
Ap.low.BP.resultParentChild
```

Enriched gene sets for genes expressed at low temperatures only.

```{r, echo=FALSE, results="asis"}
Ap.low.BP.ResTable <- GenTable(Ap.low.BP.GOdata, parentchild = Ap.low.BP.resultParentChild, topNodes=26)
Ap.low.BP.ResTable
write.table(Ap.low.BP.ResTable, file = paste(resultsdir, "Ap_low_GO.BP_results.txt", sep=""), quote=FALSE, row.names=FALSE, sep = "\t")
pandoc.table(Ap.low.BP.ResTable)
```

```{r, echo=FALSE}
# graph significant nodes
pdf(paste(resultsdir, "Ap_bimodal_BP_topGO_sig_nodes.pdf", sep=""))
showSigOfNodes(Ap.bim.BP.GOdata, score(Ap.bim.BP.resultParentChild), firstSigNodes = 52, useInfo = 'all')
dev.off()

pdf(paste(resultsdir, "Ap_intermediate_BP_topGO_sig_nodes.pdf", sep=""))
showSigOfNodes(Ap.int.BP.GOdata, score(Ap.int.BP.resultParentChild), firstSigNodes = 16, useInfo = 'all')
dev.off()

pdf(paste(resultsdir, "Ap_high_BP_topGO_sig_nodes.pdf", sep=""))
showSigOfNodes(Ap.hig.BP.GOdata, score(Ap.hig.BP.resultParentChild), firstSigNodes = 12, useInfo = 'all')
dev.off()

pdf(paste(resultsdir, "Ap_low_BP_topGO_sig_nodes.pdf", sep=""))
showSigOfNodes(Ap.low.BP.GOdata, score(Ap.low.BP.resultParentChild), firstSigNodes = 26, useInfo = 'all')
dev.off()
```

>>>>

**April 8, 2014**

@@

```{r temperature_response, eval=TRUE, echo=FALSE, results='asis'}
# table type of response
A22.type.table <- table(Ap.response.type[ , 'A22.type'])
Ar.type.table <- table(Ap.response.type[ , 'Ar.type'])
# overlap
overlap <- c(length(Ap.response.type[which(Ap.response.type$A22.type == "Bimodal" & Ap.response.type$Ar.type == "Bimodal"), "Transcript"]),
             length(Ap.response.type[which(Ap.response.type$A22.type == "High" & Ap.response.type$Ar.type == "High"), "Transcript"]),
             length(Ap.response.type[which(Ap.response.type$A22.type == "Intermediate" & Ap.response.type$Ar.type == "Intermediate"), "Transcript"]),
             length(Ap.response.type[which(Ap.response.type$A22.type == "Low" & Ap.response.type$Ar.type == "Low"), "Transcript"]),
             length(Ap.response.type[which(Ap.response.type$A22.type == "NotExp" & Ap.response.type$Ar.type == "NotExp"), "Transcript"]))

# make table
Ap.type.table <- rbind(A22.type.table, Ar.type.table, overlap)
rownames(Ap.type.table) <- c("A22", "Ar", "Both")

@@

pandoc.table(Ap.type.table, style="rmarkdown", caption = "Number of thermally-responsive transcripts with maximum expression at high, low, intermediate, both high and low (bimodal) temperatures or are not expressed for each colony and their overlap.")

@@

```{r plotthermalbreadth, echo=FALSE, eval=TRUE}
# prep data for plotting
Ap.int.sd <- rbind(A22.int.sd, Ar.int.sd)

@@

```{r GO_contingency_test}
responsive.transcripts.ann <- signif.transcripts[names(responsive.transcripts)]
# GO 'response to stress' hits in responsive transcripts
responsive.transcripts.ann[grep("GO:0006950", responsive.transcripts.ann$GO.Biological.Process), list(Transcript, best.hit.to.nr)]

resp.stress.responsive.count <- nrow(responsive.transcripts.ann[grep("GO:0006950", responsive.transcripts.ann$GO.Biological.Process), list(Transcript, best.hit.to.nr)])
# GO 'response to stress' hits in all transcripts
resp.stress.all.count <- nrow(annotationtable[grep("GO:0006950", annotationtable$GO.Biological.Process), list(Sequence.Name, best.hit.to.nr)])

GO.stress.table <- matrix(rbind(resp.stress.responsive.count, nrow(responsive.transcripts.ann) - resp.stress.responsive.count, resp.stress.all.count, nrow(annotationtable) - resp.stress.all.count), nrow = 2)

@@

There are also 6 heat shock related genes in the responsive transcripts, out of 130 total.

```{r grep_shock}
hsp_responsive <- responsive.transcripts.ann[grep("shock", responsive.transcripts.ann$best.hit.to.nr), list(Transcript, best.hit.to.nr)]
nrow(hsp_responsive)

@@

```{r gsea_all}
# create geneList. note that NA values cause problems with topGO
# so set any NA to 1 as need to retain for GO analysis
Ap.geneList <- RxNpval$padj
Ap.geneList[which(is.na(Ap.geneList))] <- 1
stopifnot(length(which(is.na(Ap.geneList))) == 0)
names(Ap.geneList) <- RxNpval$Transcript
str(Ap.geneList)

@@

# table results
Ap.BP.ResTable <- GenTable(Ap.BP.GOdata, parentchild = Ap.BP.resultParentChild, topNodes = 82)
#Ap.BP.ResTable
write.table(Ap.BP.ResTable, file = paste(resultsdir, "Ap_GO.BP_results.txt", sep=""), quote=FALSE, row.names=FALSE, sep = "\t")
pandoc.table(Ap.BP.ResTable)

@@

**2) High**

Genes with *High* expression in both colonies

Use `selectTranscript` function to select transcripts from 'Ap.response.type' for GSEA.

```{r gsea_high}
# Select transcripts
Ap.high <- Ap.response.type[which(Ap.response.type$Ar.type == "High" & Ap.response.type$A22.type == "High"), "Transcript"]
Ap.geneList.high <- rep(0, length = length(Ap.geneList))
names(Ap.geneList.high) <- names(Ap.geneList)
Ap.geneList.high[(which(names(Ap.geneList.high) %in% Ap.high))] <- 1
# check correct number of values set to 1
table(Ap.geneList.high)

# Run GSEA
Ap.high.gsea <- gsea(genelist = Ap.geneList.high, geneID2GO = geneID2GO)
pandoc.table(Ap.high.gsea)
```

**3) Low**

Genes with *Low* expression in both colonies

```{r gsea_low}
# Select transcripts
Ap.low <- Ap.response.type[which(Ap.response.type$Ar.type == "Low" & Ap.response.type$A22.type == "Low"), "Transcript"]
Ap.geneList.low <- rep(0, length = length(Ap.geneList))
names(Ap.geneList.low) <- names(Ap.geneList)
Ap.geneList.low[(which(names(Ap.geneList.low) %in% Ap.low))] <- 1
# check correct number of values set to 1
table(Ap.geneList.low)

# Run GSEA
Ap.low.gsea <- gsea(genelist = Ap.geneList.low, geneID2GO = geneID2GO)
pandoc.table(Ap.low.BP.ResTable)
```

**4) Bimodal**

Genes with *Bimodal* expression in both colonies

```{r gsea_bimodal}
Ap.bim <- Ap.response.type[which(Ap.response.type$A22.type == "Bimodal" & Ap.response.type$Ar.type == "Bimodal"), "Transcript"]
# create gene list, setting value to 1 for "bim" transcripts
Ap.geneList.bim <- rep(0, length = length(Ap.geneList))
names(Ap.geneList.bim) <- names(Ap.geneList)
Ap.geneList.bim[(which(names(Ap.geneList.bim) %in% Ap.bim))] <- 1
# check correct number of values set to 1
table(Ap.geneList.bim)

# Run GSEA
Ap.bim.gsea <- gsea(genelist = Ap.geneList.bim, geneID2GO = geneID2GO)
pandoc.table(Ap.bim.gsea)
```

**5) Intermediate**

Genes with *Intermediate* expression in both colonies

```{r gsea_intermediate}
Ap.int <- Ap.response.type[which(Ap.response.type$A22.type == "Intermediate" & Ap.response.type$Ar.type == "Intermediate"), "Transcript"]
# create gene list, setting value to 1 for "int" transcripts
Ap.geneList.int <- rep(0, length = length(Ap.geneList))
names(Ap.geneList.int) <- names(Ap.geneList)
Ap.geneList.int[(which(names(Ap.geneList.int) %in% Ap.int))] <- 1
# check correct number of values set to 1
table(Ap.geneList.int)

# Run GSEA
Ap.int.gsea <- gsea(genelist = Ap.geneList.int, geneID2GO = geneID2GO)
pandoc.table(Ap.int.gsea)
```

Next, I perform GSEA for genes in each functional type in one colony but not the other (e.g. the set difference) to gain insight on differences between the colonies.

**A22 'High' genes not in Ar**

```{r gsea_high_A22}
A22.high <- Ap.response.type[which(Ap.response.type$A22.type == "High" & Ap.response.type$Ar.type != "High"), "Transcript"]
# create gene list, setting value to 1 for "bim" transcripts
A22.geneList.high <- rep(0, length = length(Ap.geneList))
names(A22.geneList.high) <- names(Ap.geneList)
A22.geneList.high[(which(names(A22.geneList.high) %in% A22.high))] <- 1
# check correct number of values set to 1
table(A22.geneList.high)

# Run GSEA
A22.high.gsea <- gsea(genelist = A22.geneList.high, geneID2GO = geneID2GO)
pandoc.table(A22.high.gsea)
```

**Ar 'High' genes not in Ar**

```{r gsea_high_Ar}
Ar.high <- Ap.response.type[which(Ap.response.type$A22.type != "High" & Ap.response.type$Ar.type == "High"), "Transcript"]
# create gene list, setting value to 1 for "bim" transcripts
Ar.geneList.high <- rep(0, length = length(Ap.geneList))
names(Ar.geneList.high) <- names(Ap.geneList)
Ar.geneList.high[(which(names(Ar.geneList.high) %in% Ar.high))] <- 1
# check correct number of values set to 1
table(Ar.geneList.high)

# Run GSEA
Ar.high.gsea <- gsea(genelist = Ar.geneList.high, geneID2GO = geneID2GO)
pandoc.table(Ar.high.gsea)
```

**A22 'Low' genes not in Ar**

```{r gsea_low_A22}
A22.low <- Ap.response.type[which(Ap.response.type$A22.type == "Low" & Ap.response.type$Ar.type != "Low"), "Transcript"]
# create gene list, setting value to 1 for "bim" transcripts
A22.geneList.low <- rep(0, length = length(Ap.geneList))
names(A22.geneList.low) <- names(Ap.geneList)
A22.geneList.low[(which(names(A22.geneList.low) %in% A22.low))] <- 1
# check correct number of values set to 1
table(A22.geneList.low)

# Run GSEA
A22.low.gsea <- gsea(genelist = A22.geneList.low, geneID2GO = geneID2GO)
pandoc.table(A22.low.gsea)
```

**Ar 'Low' genes not in Ar**

```{r gsea_low_Ar}
Ar.low <- Ap.response.type[which(Ap.response.type$A22.type != "Low" & Ap.response.type$Ar.type == "Low"), "Transcript"]
# create gene list, setting value to 1 for "bim" transcripts
Ar.geneList.low <- rep(0, length = length(Ap.geneList))
names(Ar.geneList.low) <- names(Ap.geneList)
Ar.geneList.low[(which(names(Ar.geneList.low) %in% Ar.low))] <- 1
# check correct number of values set to 1
table(Ar.geneList.low)

# Run GSEA
Ar.low.gsea <- gsea(genelist = Ar.geneList.low, geneID2GO = geneID2GO)
pandoc.table(Ar.low.gsea)
```

**A22 'Bimodal' genes not in Ar**

```{r gsea_bim_A22}
A22.bim <- Ap.response.type[which(Ap.response.type$A22.type == "Bimodal" & Ap.response.type$Ar.type != "Bimodal"), "Transcript"]
# create gene list, setting value to 1 for "bim" transcripts
A22.geneList.bim <- rep(0, length = length(Ap.geneList))
names(A22.geneList.bim) <- names(Ap.geneList)
A22.geneList.bim[(which(names(A22.geneList.bim) %in% A22.bim))] <- 1
# check correct number of values set to 1
table(A22.geneList.bim)

# Run GSEA
A22.bim.gsea <- gsea(genelist = A22.geneList.bim, geneID2GO = geneID2GO)
pandoc.table(A22.bim.gsea)
```

**Ar 'Bimodal' genes not in Ar**

```{r gsea_bim_Ar}
Ar.bim <- Ap.response.type[which(Ap.response.type$A22.type != "Bimodal" & Ap.response.type$Ar.type == "Bimodal"), "Transcript"]
# create gene list, setting value to 1 for "bim" transcripts
Ar.geneList.bim <- rep(0, length = length(Ap.geneList))
names(Ar.geneList.bim) <- names(Ap.geneList)
Ar.geneList.bim[(which(names(Ar.geneList.bim) %in% Ar.bim))] <- 1
# check correct number of values set to 1
table(Ar.geneList.bim)

# Run GSEA
Ar.bim.gsea <- gsea(genelist = Ar.geneList.bim, geneID2GO = geneID2GO)
pandoc.table(Ar.bim.gsea)
```

**A22 'Intermediate' genes not in Ar**

```{r gsea_int_A22}
A22.int <- Ap.response.type[which(Ap.response.type$A22.type == "Intermediate" & Ap.response.type$Ar.type != "Intermediate"), "Transcript"]
# create gene list, setting value to 1 for "bim" transcripts
A22.geneList.int <- rep(0, length = length(Ap.geneList))
names(A22.geneList.int) <- names(Ap.geneList)
A22.geneList.int[(which(names(A22.geneList.int) %in% A22.int))] <- 1
# check correct number of values set to 1
table(A22.geneList.int)

# Run GSEA
A22.int.gsea <- gsea(genelist = A22.geneList.int, geneID2GO = geneID2GO)
pandoc.table(A22.int.gsea)
```

**Ar 'Low' genes not in Ar**

```{r gsea_int_Ar}
Ar.int <- Ap.response.type[which(Ap.response.type$A22.type != "Intermediate" & Ap.response.type$Ar.type == "Intermediate"), "Transcript"]
# create gene list, setting value to 1 for "bim" transcripts
Ar.geneList.int <- rep(0, length = length(Ap.geneList))
names(Ar.geneList.int) <- names(Ap.geneList)
Ar.geneList.int[(which(names(Ar.geneList.int) %in% Ar.int))] <- 1
# check correct number of values set to 1
table(Ar.geneList.int)

# Run GSEA
Ar.int.gsea <- gsea(genelist = Ar.geneList.int, geneID2GO = geneID2GO, plotpath = "results/Ar_int_BP_topGO_sig_nodes.pdf")
pandoc.table(Ar.int.gsea)
```
